---
title: "Creating your first ggrank plot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating your first ggrank plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 10, fig.height = 6)
```

`ggrank` shows how categories move through an ordered ranking while retaining
their underlying values. Start with one row per category and period and map the
three required columns.

> **Teaching-data notice:** All datasets bundled with `ggrank` are synthetic.
> They do not contain Global Burden of Disease estimates. `ggrank` is an
> independent project and is not affiliated with or endorsed by the Institute
> for Health Metrics and Evaluation (IHME).

```{r}
library(ggrank)

ggrank(
  ggrank_products,
  category = product,
  period = year,
  value = sales,
  periods = c(2022, 2024),
  top_n = 5,
  value_header = "Sales"
)
```

The boundary view is the default: a category outside the top five remains in
the figure when it enters or exits the top five in another displayed period.

## Group colours and prepared labels

Use `group` for meaningful category colours and `label` when values require a
domain-specific display format. Supplied ranks are also supported.

```{r}
ggrank(
  ggrank_causes,
  category = cause,
  period = year,
  value = rate,
  rank = rank,
  label = display_value,
  group = cause_group,
  periods = c(1990, 2021),
  top_n = 10,
  value_header = "Rate (95% interval)"
)
```

The returned value is a regular ggplot object, so titles, captions, and other
ggplot2 layers can be added normally.

## Inspect the underlying comparison

`ggrank_table()` returns a readable analytical companion with one row per
category and adjacent transition.

```{r}
ggrank_table(
  ggrank_products,
  category = product,
  period = year,
  value = sales,
  periods = c(2022, 2024),
  top_n = 5
)
```
