---
title: "Rank changes, movement status, and legends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rank changes, movement status, and legends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 6)
```

`ggrank` has two primary visual questions:

- `ggrank()` asks **How did the ranking evolve?**
- `ggrank_change()` asks **Who moved the most?**

Both use the same ranks. The change chart is not a second ranking algorithm.

## How rank change is calculated

For each adjacent comparison:

```text
rank_change = rank_from - rank_to
```

Rank 7 to rank 3 is `+4`: the category rose four positions. Rank 2 to rank 5
is `-3`: it fell three positions. Rank 3 to rank 3 is zero. Competition ties
remain statistical ties; `display_position` only prevents tied categories from
overlapping in `ggrank()`.

```{r}
library(ggrank)

changes <- ggrank_table(
  ggrank_products, product, year, sales,
  top_n = 5,
  show_transitions = "all"
)

changes[c("category", "from", "to", "rank_from", "rank_to",
          "rank_change", "status")]
```

## What each status means

`ggrank_table()` preserves boundary and data-availability information:

| Status | Meaning |
|---|---|
| `riser` | Rank improved and the category did not cross into the selected top N. |
| `faller` | Rank worsened and the category did not cross out of the selected top N. |
| `stable` | Statistical rank did not change. |
| `entrant` | Moved from outside the top-N boundary to inside it. |
| `exit` | Moved from inside the top-N boundary to outside it. |
| `new` | No earlier-period rank is available. |
| `absent` | No later-period rank is available. |
| `missing` | A non-finite value was supplied for either side. |

The `ggrank_change()` colour is intentionally simpler. It uses the sign of
`rank_change`: an entrant moving 6 to 4 is shown as a riser; an exit moving 3
to 7 is shown as a faller. The original `status` remains in the plot data.

## Latest, all, and explicit comparisons

The default focuses on the latest consecutive comparison and removes stable
categories:

```{r}
ggrank_change(changes, top = 5)
```

`top = 5` means the five largest absolute rank changes in the selected
comparison—not five risers plus five fallers.

```{r}
ggrank_change(changes, top = 5, comparison = "all")
```

Select a comparison already represented in the table with `from` and `to`:

```{r}
ggrank_change(changes, from = 2022, to = 2023, top = 5)
```

Stable categories are optional:

```{r}
ggrank_change(changes, top = 8, show_stable = TRUE)
```

## Detailed labels and long names

```{r}
ggrank_change(
  changes,
  top = 5,
  change_label = "ranks",
  label_wrap = 24
)
```

Use `change_label = "change"` for `+4` and `-3`, `"ranks"` for
`7 → 3 (+4)`, or `"none"` for no bar-end annotation.

## Customise movement colours and legend text

```{r}
movement_colours <- c(
  riser = "#0072B2",
  faller = "#D55E00",
  stable = "#667085"
)

movement_labels <- c(
  riser = "Moved towards rank 1",
  faller = "Moved away from rank 1",
  stable = "No rank change"
)

ggrank_change(
  changes,
  top = 5,
  palette = movement_colours,
  legend_title = "Movement",
  legend_labels = movement_labels
)
```

Set `show_legend = FALSE` when direction, labels, and explanatory text make the
legend unnecessary. Because the result is a ggplot, standard scale and theme
functions remain available for further refinement.

## Customise group legends in `ggrank()`

When `group` is supplied, `colour_by = "auto"` uses its values. Palette and
legend-label names must match those values exactly.

```{r}
cause_colours <- c(
  "Communicable" = "#009E73",
  "Injuries" = "#0072B2",
  "Non-communicable" = "#D55E00"
)

cause_labels <- c(
  "Communicable" = "Communicable diseases",
  "Injuries" = "Injuries",
  "Non-communicable" = "Non-communicable diseases"
)

ggrank(
  ggrank_causes, cause, year, rate,
  rank = rank, label = display_value, group = cause_group,
  periods = c(1990, 2021), top_n = 10,
  palette = cause_colours,
  legend_title = "Cause group",
  legend_labels = cause_labels
)
```

To use movement rather than group colours, choose
`colour_by = "movement"`. To hide the legend, use `show_legend = FALSE`.
