---
title: "Sports leaderboards and larger rankings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sports leaderboards and larger rankings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 12, fig.height = 8)
```

Rankings are not limited to health data. `ggrank` can show football tables,
cricket standings, gaming ladders, university rankings, sales leaderboards,
and other ordered lists.

All names and results in this article are synthetic teaching data. They are not
official English Premier League, UEFA Champions League, Indian Premier League,
or other competition results, and the package is not affiliated with those
organisations.

## What does `top_n` mean?

`top_n` is the highest rank included in the main display boundary:

* `top_n = 5` focuses on the top five;
* `top_n = 10` focuses on the top ten;
* `top_n = 20` can display a complete twenty-club league;
* `top_n = 30` can display thirty ranks.

It is deliberately a rank threshold rather than an exact row count. With
competition ranking, everyone tied at the boundary is retained. A top-10
figure can therefore contain eleven or more rows.

## A twenty-club football table

This Premier League-style example supplies points and lets `ggrank()` calculate
the ranks separately within each season.

```{r football-data}
library(ggrank)

football <- data.frame(
  season = rep(c("2024", "2025"), each = 20),
  club = rep(c(
    "Ashford Athletic", "Bayside City", "Cedar United", "Docklands FC",
    "Eastborough", "Forest Rovers", "Granite Town", "Harbour Albion",
    "Ironbridge", "Juniper Wanderers", "Kingsport", "Lakeside FC",
    "Meadow County", "Northgate", "Oakfield", "Parkside Rangers",
    "Queensborough", "Riverside", "Stonehaven", "Westford"
  ), 2),
  points = c(
    91, 86, 82, 78, 74, 70, 66, 63, 60, 57,
    54, 51, 48, 45, 42, 39, 36, 33, 29, 24,
    84, 92, 76, 81, 69, 73, 61, 65, 58, 55,
    52, 49, 46, 43, 40, 37, 34, 31, 27, 22
  )
)
```

A top-five view is compact and focuses on title contenders:

```{r football-top-five, fig.height=5}
ggrank(
  football, club, season, points,
  top_n = 5,
  value_header = "Points",
  title = "Football title race"
)
```

Use `top_n = 20` for the complete league table:

```{r football-full, fig.height=11}
football_plot <- ggrank(
  football, club, season, points,
  top_n = 20,
  value_header = "Points",
  label_wrap = 20,
  base_size = 9,
  title = "Complete football league table"
)

football_plot
```

## A ten-team cricket-style leaderboard

When official ranks already exist, points or net run rate are optional. This
IPL-style teaching example therefore uses names and ranks only.

```{r cricket, fig.height=7}
cricket <- data.frame(
  season = rep(c("2025", "2026"), each = 10),
  team = rep(c(
    "Bengaluru Blazers", "Chennai Waves", "Delhi Comets", "Gujarat Falcons",
    "Hyderabad Suns", "Jaipur Royals", "Kolkata Knights", "Lucknow Leopards",
    "Mumbai Mariners", "Punjab Panthers"
  ), 2),
  rank = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10,
           4, 1, 5, 2, 8, 3, 6, 10, 7, 9)
)

ggrank(
  cricket, team, season,
  rank = rank,
  top_n = 10,
  label_wrap = 20,
  title = "Cricket league standings"
)
```

The rank-only layout omits the value column automatically.

## Choosing an export height

The package does not impose a hard row limit, but a fixed-size page does. These
are practical starting points for a two-period chart:

| Displayed rows | Suggested height | Typical use |
|---:|---:|---|
| 5 | 5–7 inches | slide or compact report |
| 10 | 7–9 inches | presentation or landscape page |
| 20 | 10–12 inches | full league table or tall page |
| 30 | 14–16 inches | poster, scrolling report, or large-format export |

```{r export, eval=FALSE}
ggplot2::ggsave(
  "football-table.png", football_plot,
  width = 14, height = 11, units = "in", dpi = 300
)
```

Long names, wrapped labels, four periods, and larger text require additional
height or width. For a normal presentation slide, showing the top five or top
ten is usually clearer than squeezing thirty rows into one panel. A complete
top-30 figure is better suited to a tall image, poster, webpage, or appendix.
